---
title: "Note on potential discrepancies between simtrial and survdiff"
author: "Larry Leon"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Note on potential discrepancies between simtrial and survdiff}
  %\VignetteEngine{knitr::rmarkdown}
---

## Overview

```{r, message=FALSE}
library(gsDesign)
library(gsDesign2)
library(dplyr)
library(tibble)
library(gt)
#library(ggplot2)
#library(cowplot)
library(simtrial)
library(tidyr)
#library(future.batchtools)
#library(doFuture)
#library(foreach)

#library(tictoc)

library(survival)


```

In the **survival** (base R) package the log-rank and Cox estimation procedures apply (by default) a correction to "fix" roundoff errors.   These are implemented with the *timefix* option (by default *timefix = TRUE*) via the *aeqSurv* function.   
However in the **simtrial** package (and also **Hmisc**) such a correction is not implemented;  Consequently, there can be discrepancies between **simtrial** and base R *survival* (*survdiff*, *coxph*, and *survfit*).

For details on the *aeqSurv* function see [Therneau, 2016](https://cran.r-project.org/web/packages/survival/vignettes/tiedtimes.pdf) and [R documentation, version 3.803](https://www.rdocumentation.org/packages/survival/versions/3.8-3/topics/aeqSurv)

In the following we describe a simulation scenario where a discrepancy is generated and illustrate how discrepancies can be resolved (if desired) by pre-processing survival times with *aeqSurv* and thus replicating *survdiff* and *coxph* default calculations.

In the simulated dataset two observations are generated:

-   Observation $i=464$ with survival time $Y=0.306132722582$
-   Observation $i=516$ with survival time $Y=0.306132604679$
-   Per "aeqSurv" these times are tied and set to $Y=0.306132604679$
-   The log-rank and Cox estimates can therefore differ between other approaches without the "timefix" correction


## Scenario definitions

We define various true data generating model scenarios and convert for use in **gsDesign2**.  Here we are using a single scenario where discrepancies were found.   This is just for illustration to inform the user of **simtrial** that discrepancies can occur and how to resolve via *aeqSurv*, if desired.


```{r}
survival_at_24_months <- 0.35
hr <- log(.35)/log(.25)
control_median <- 12
control_rate <- c(log(2) / control_median, (log(.25) - log(.2)) / 12)

scenarios <- tribble(
  ~Scenario, ~Name,           ~Period, ~duration, ~Survival,
  0,         "Control",       0,       0,         1,
  0,         "Control",       1,       24,        .25,
  0,         "Control",       2,       12,        .2,
  1,         "PH",            0,       0,         1,
  1,         "PH",            1,       24,        .35,
  1,         "PH",            2,       12,        .2^hr,
  2,         "3-month delay", 0,       0,         1,
  2,         "3-month delay", 1,       3,         exp(-3 * control_rate[1]),
  2,         "3-month delay", 2,       21,        .35,
  2,         "3-month delay", 3,       12,        .2^hr,
  3,         "6-month delay", 0,       0,         1,
  3,         "6-month delay", 1,       6,         exp(-6 * control_rate[1]),
  3,         "6-month delay", 2,       18,        .35,
  3,         "6-month delay", 3,       12,        .2^hr,
  4,         "Crossing",      0,       0,         1,
  4,         "Crossing",      1,       3,         exp(-3 * control_rate[1] * 1.3),
  4,         "Crossing",      2,       21,        .35,
  4,         "Crossing",      3,       12,        .2^hr,
  5,         "Weak null",     0,       0,         1,
  5,         "Weak null",     1,       24,        .25,
  5,         "Weak null",     2,       12,        .2,
  6,         "Strong null",   0,       0,         1,         
  6,         "Strong null",   1,       3,         exp(-3 * control_rate[1] * 1.5),
  6,         "Strong null",   2,       3,         exp(-6 * control_rate[1]),
  6,         "Strong null",   3,       18,        .25,
  6,         "Strong null",   4,       12,        .2,
  )
# scenarios |> gt()
```


```{r}
fr <- 
  scenarios |>
    group_by(Scenario) |>
  #  filter(Scenario == 2) |>
    mutate(Month = cumsum(duration),
           x_rate = -(log(Survival) - log(lag(Survival, default = 1))) / 
           duration, 
           rate = ifelse(Month > 24, control_rate[2], control_rate[1]),
           hr = x_rate / rate) |>
    select(-x_rate) |>
    filter(Period > 0, Scenario > 0) |> ungroup()
#fr |> gt() |> fmt_number(columns = everything(), decimals = 2)

fr <- fr |> mutate(fail_rate = rate, dropout_rate =0.001, stratum = "All")


# MWLR
mwlr <- fixed_design_mb(
  tau = 12,
  enroll_rate = define_enroll_rate(duration = 12, rate = 1),
  fail_rate = fr |> filter(Scenario == 2),
  alpha = 0.025, power = .85, ratio = 1,
  study_duration = 36
) |> to_integer()


er <- mwlr$enroll_rate

```

# A scenario that generates a discrepancy 


```{r}

set.seed(3219)

dgm <- fr[c(14:17),]

fail_rate <- data.frame(stratum = rep("All", 2 * nrow(dgm)),
                          period = rep(dgm$Period, 2),
                          treatment = c(rep("control", nrow(dgm)), 
                                        rep("experimental", nrow(dgm))),
                          duration = rep(dgm$duration, 2),
                          rate = c(dgm$rate, dgm$rate * dgm$hr)
                         )
    
dgm$stratum <- "All"
# Constant dropout rate for both treatment arms and all scenarios
dropout_rate <- data.frame(stratum = rep("All", 2), 
                           period = rep(1, 2),
                           treatment = c("control", "experimental"),
                           duration = rep(100, 2),
                           rate = rep(.001, 2)
                          )
```


Simulated dataset with discrepancy between logrank test of *wlr* (**simtrial**) and *survdiff* (also compare to score test of *coxph* [same as survdiff with default *timefix=TRUE*])

```{r}

ss <- 395

set.seed(8316951+ss*1000)  
  
# Generate a dataset
dat <- sim_pw_surv(n = 698, enroll_rate = er,
fail_rate = fail_rate, dropout_rate = dropout_rate)

analysis_data <- cut_data_by_date(dat, 36)

dfa <- analysis_data

dfa$treat <- ifelse(dfa$treatment=="experimental",1,0)

z1 <- dfa |>  wlr(weight=fh(rho=0,gamma=0))

check <- survdiff(Surv(tte,event)~ treat, data=dfa)

# Note, for coxph use 
#cph.score <- summary(coxph(Surv(tte,event)~ treat, data=dfa, control=coxph.control(timefix=TRUE)))$sctest

cat("Log-rank wlr() vs survdiff()",c(z1$z^2,check$chisq),"\n")
```

Verify that *timefix=FALSE* in *coxph* agrees with *wlr* 

```{r}
cph.score <- summary(coxph(Surv(tte,event)~ treat, data=dfa, control=coxph.control(timefix=FALSE)))$sctest
cat("Log-rank wlr() vs Cox score z^2",c(z1$z^2,cph.score["test"]),"\n")
```


Pre-processing survival times with *aeqSurv* to implement *timefix=TRUE* procedure.  

Verify *wlr* and *survdiff* now agree.

```{r}
Y <- dfa[,"tte"]
Delta <- dfa[,"event"]

tfixed <- aeqSurv(Surv(Y,Delta))
Y<- tfixed[,"time"]
Delta <- tfixed[,"status"]
# Use aeqSurv version
dfa$tte2 <- Y
dfa$event2 <- Delta

# wlr() after "timefix"
dfa2 <- dfa
dfa2$tte <- dfa2$tte2
dfa2$event <- dfa2$event2
z1new <- dfa2 |>  wlr(weight=fh(rho=0,gamma=0))
cat("Log-rank wlr() with timefix vs survdiff() z^2",c(z1new$z^2,check$chisq),"\n")

```


Where do they differ (tte2 are times after *aeqSurv*) ?

```{r}

dfa <- dfa[order(dfa$tte2),]

id <- seq(1,nrow(dfa))

diff <- exp(dfa$tte) - exp(dfa$tte2)
id_diff <- which(abs(diff)>0)

tolook <- seq(id_diff-2,id_diff+2)

dfcheck <-  dfa[tolook,c("tte","tte2","event","event2","treatment")]
print(dfcheck,digits=12)
```




Verify *coxph* (default) and *coxph* with aeqSurv pre-processing (using tte2 as outcome and setting *timefix=FALSE*) are identical:

Also note that here ties do not have impact because in separate arms

```{r}
# Check Cox with ties
cox_breslow <- summary(coxph(Surv(tte,event)~treatment,data=dfa,ties="breslow"))$conf.int
cox_efron <- summary(coxph(Surv(tte,event)~treatment,data=dfa,ties="efron"))$conf.int
cat("Cox Breslow and Efron hr (tte, timefix=TRUE):",c(cox_breslow[1],cox_efron[1]),"\n")

# Here ties do not have impact because in separate arms
cox_breslow <- summary(coxph(Surv(tte2,event2)~treatment,data=dfa,ties="breslow", control=coxph.control(timefix=FALSE)))$conf.int
cox_efron <- summary(coxph(Surv(tte2,event2)~treatment,data=dfa,ties="efron", control=coxph.control(timefix=FALSE)))$conf.int
cat("Cox Breslow and Efron hr (tte2, timefix=FALSE):",c(cox_breslow[1],cox_efron[1]),"\n")
```

**So here there is a difference between tte and tte2 times, but there is not an impact of ties for Cox between *breslow* and *efron* because the ties (single tie in tte2) are in separate arms**.

Lastly, artificially change treatment so that two observations are tied within the same treatment arm which generates difference between *breslow* and *efron* options for ties:


```{r}
# Create tie within treatment arm by changing treatment 
dfa3 <- dfa
dfa3[19,"treat"]  <- 1.0

cox_breslow <- summary(coxph(Surv(tte,event)~treat, data=dfa3,ties="breslow", control=coxph.control(timefix=TRUE)))$conf.int
cox_efron <- summary(coxph(Surv(tte,event)~treat, data=dfa3,ties="efron", control=coxph.control(timefix=TRUE)))$conf.int
cat("Cox Breslow and Efron hr (tte, timefix=TRUE)=",c(cox_breslow[1],cox_efron[1]),"\n")

```


Same as

```{r}

cox_breslow <- summary(coxph(Surv(tte2,event2)~treat, data=dfa3,ties="breslow", control=coxph.control(timefix=FALSE)))$conf.int
cox_efron <- summary(coxph(Surv(tte2,event2)~treat, data=dfa3,ties="efron", control=coxph.control(timefix=FALSE)))$conf.int
cat("Cox Breslow and Efron hr (tte2, timefix=FALSE)=",c(cox_breslow[1],cox_efron[1]),"\n")

```

